\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-17 NAR_DRAFT.tex
hate the figure placement hassle
get there eventually
good figures now
number all of the equations for consistency

\rngcomment{Got here...}
\bkmcomment{No currently published paper has numbered equations, and it is not mentioned in the guidelines. I am using their template for numbered and non-numbered equations.}
We model the observed mutation profile $S$ (Figure~\ref{NAR-sigfig}\textbf{A} throught \textbf{C}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model $\mathrm{p}(\mathrm{S})$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
$$
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
$$
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 with the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
The R implementation for \batcave ships with pre-computed tables for both whole exomes and whole genomes.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/ppv_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points}
\label{NAR-ppv_fig}
\end{figure}

>>>>

\rngcomment{Got here...}
\bkmcomment{No currently published paper has numbered equations, and it is not mentioned in the guidelines. I am using their template for numbered and non-numbered equations.}
We model the observed mutation profile $S$ (Figure~\ref{NAR-sigfig}\textbf{A} throught \textbf{C}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model $\mathrm{p}(\mathrm{S})$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 with the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
\begin{equation}
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
\end{equation}

Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
The R implementation for \batcave ships with pre-computed tables for both whole exomes and whole genomes.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points}
\label{NAR-ppv_fig}
\end{figure}
